### 2017-08-10 doc/MSc-dissertation.tex
DNS almost done

\usepackage{graphics}
\usepackage{fullpage,epsf,graphicx, amstext,url,bm,amsmath, amssymb}

@@

@@

\section{DNS basics}

@@

The main problem with computational fluid dynamics is the nonlinear term $(\bm u \cdot \bm \nabla) \bm u$ in the Navier-Stokes (and MHD) equations. As it was mentioned previously, this term is directly related to turbulence in a system. In order to reformulate the Navier-Stokes equations to have only one dynamic variable, the velocity field, it is necessary to Fourier transform (FT) them and to use the incompressibility condition $\dot{p}=\bm \nabla \cdot \bm u = 0$ (or an equation of state $p=p(\rho)$)\footnote{For MHD two dynamic variables will exist, the velocity and the magnetic fields.}. This means to set the pressure term as a function of the velocity field. However, the nonlinear term remains in the new equations and now it is a convolution over all possible wave numbers~\cite{mccomb1990physics}!

After generating a velocity field, the Fourier-transformed Navier-Stokes equations are then numerically solved. It is worth noting that the nonlinear convolution alone will require $\mathcal{O}(N^6)$ operations~\cite{mccomb1990physics}. This will scale horribly when dealing with large boxes. But wait, that's $\mathcal{O}(N^6)$ operations for the nonlinear term per time step! In order to mitigate this problem, the convolution theorem is used. The convolution theorem states that a convolution's Fourier transform $\mathcal{F}\{f \ast g\}$ is defined by the point-wise product of the Fourier transform of its components $\mathcal{F}\{f\} \cdot \mathcal{F}\{g\}$. This means that the convolution can be described as $f \ast g = \mathcal{F}^{-1}\{ \mathcal{F}\{f\} \cdot \mathcal{F}\{g\} \}$. Using fast Fourier transforms (FFT), and considering that for the nonlinear term in the Navier-Stokes equation both $f$ and $g$ are the velocity field $\bm u (\bm k, t)$, the computational complexity for obtaining the nonlinear term is decreased to $\mathcal{O}(N^3 \log_2 N)$~\cite{mccomb1990physics}.

@@

Afterwards, the velocity and magnetic fields are

@@

%\appendix
% the appendix command just changes heading styles for appendices.

%\chapter{Stuff that won't be read by anyone}

%Some people include in their thesis a lot of detail, particularly lots of tables containing raw results, figures of intermediate results, or computer code which no-one will ever read. You should be careful that anything like this you include should contain some element of uniqueness which justifies its inclusion.

>>>>

\usepackage{graphics}
\usepackage{listings}
\usepackage{fullpage,epsf,graphicx, amstext,url,bm,amsmath, amssymb}

@@

\lstset{breaklines=true}

@@

\section{DNS basics}
\label{sec4.1}

@@

The main problem with computational fluid dynamics is the nonlinear term $(\bm u \cdot \bm \nabla) \bm u$ in the Navier-Stokes (and MHD) equations. As it was mentioned previously, this term is directly related to turbulence in a system. In order to reformulate the Navier-Stokes equations to have only one dynamic variable\footnote{For MHD two dynamic variables will exist, the velocity and the magnetic fields.}, the velocity field, it is necessary to Fourier transform (FT) them and to use the incompressibility condition $\dot{p}=\bm \nabla \cdot \bm u = 0$ (or an equation of state $p=p(\rho)$). This means to set the pressure term as a function of the velocity field. However, the nonlinear term remains in the new equations and now it is a convolution over all possible wave numbers~\cite{mccomb1990physics}!

After generating a velocity field, the Fourier-transformed Navier-Stokes equations are then numerically solved. It is worth noting that the nonlinear convolution alone will require $\mathcal{O}(N^6)$ operations~\cite{mccomb1990physics}. This will scale horribly when dealing with large boxes. But wait, that's $\mathcal{O}(N^6)$ operations for the nonlinear term per time step! In order to mitigate this problem, the convolution theorem is used. The convolution theorem states that a convolution's Fourier transform $\mathcal{F}\{f \ast g\}$ is defined by the point-wise product of the Fourier transform of its components $\mathcal{F}\{f\} \cdot \mathcal{F}\{g\}$. This means that the convolution can be described as $f \ast g = \mathcal{F}^{-1}\{ \mathcal{F}\{f\} \cdot \mathcal{F}\{g\} \}$. Using fast Fourier transforms\footnote{Since Fourier transforms involve infinite sums, which cannot be numerically performed, FFTs are used as an alternative. This, however, generates aliasing errors. More on aliasing errors on eDNS2017 can be found in ~\cite{yoffe2013investigation}.} (FFT), and considering that for the nonlinear term in the Navier-Stokes equation, the computational complexity for obtaining the nonlinear term is decreased to $\mathcal{O}(N^3 \log_2 N)$~\cite{mccomb1990physics}.

@@

Afterwards, the velocity and magnetic fields are initialised. This is done by using Gaussian distributions (with mean 0 and variance 1) for the velocity and magnetic fields in configuration space~\cite{yoffe2013investigation}. These fields do not have to be a solution of the MHD equations. These are then Fourier transformed to inject initial helicity into the system (depending on the given initial conditions). Both fields are then normalised with an equipartition of kinetic and magnetic energy.

Fourier analysis methods are widely used when systems are described by coupled partial differential equations. This is mainly because only temporal derivatives remain and spatial derivatives transform as $\bm \nabla \rightarrow -i \bm k$. This turns partial differential equations to linear ordinary differential equations in Fourier space. As it was mentioned before, the convolution term when dealing with the Fourier representation of the Navier-Stokes (or MHD) equations is problematic and can be dealt with by using the convolution theorem. To do so, FFTs are used to go back and forth between spectral and configuration space. The time stepping strategy used for eDNS2017 is Heun's method, which is an explicit second-order predictor-corrector scheme. A more detailed description of the eDNS2017 code can be found in~\cite{LinkmannMoritzFrederikLeon2016Spim, yoffe2013investigation}.

\subsection{Forcing}

Another important part of the eDNS2017 code is the type of forcing used to inject energy into the system. It is worth noting that the code allows to study isotropic turbulence for decaying (no energy injected into the system) or forcing (energy injected in order to reach a steady state)~\cite{LinkmannMoritzFrederikLeon2016Spim}. This forcing comes as $\bm f_u$ and $\bm f_b$ in eqns.~\ref{eq3.1} and~\ref{eq3.2}. However, for the simulations that were performed only the velocity field was forced, and from it energy was transferred to the magnetic field through the Lorentz force. From stationary simulations, the forcing function used injects energy by feeding a rescaling of the velocity field into the large spatial scales. More precisely, $\bm f_u$ is given by:

\begin{align}
 \bm f_u(\bm k, t) &= \frac{\varepsilon_W}{2 E_f} \bm u(\bm k, t) \qquad \text{for} \qquad 0 < \vert \bm k \vert \leq k_f; \nonumber \\
 &= 0 \,\, \qquad \qquad \qquad \text{otherwise,} \nonumber
\end{align}

where $\varepsilon_W$ is the energy input rate, $E_f$ is the total energy contained in the forcing wave number band ($0 < \vert \bm k \vert \leq k_f$), and $k_f$ is the largest wave number for which forcing is applied ($k_f=2$ in all simulations). Finally, $\bm u(\bm k, t)$ is the velocity field in spectral space, i.e. the Fourier transform of the velocity field in configuration space~\cite{LinkmannMoritzFrederikLeon2016Spim}. This can be defined as an input parameter in the eDNS2017 code.

Determining the value of the energy input rate needs to be carefully considered. The main considerations that need to be made to determine what numerical value to assign to $\varepsilon_W$ are the time step $dt$, and the energy dissipation of the system. In order for a forced simulation to obtain a steady state, it must ensure that $\frac{dE}{dt}=\varepsilon_W - \varepsilon_T = 0$, where $\varepsilon_T$ is the total energy dissipation.

\subsubsection{Helical forcing}

In order to achieve large-scale dynamo action, the system must have kinetic helicity. The best way to do this is to have the forcing routine inject helicity to the system. This can be done by using a forcing routine given by:

\begin{align}
 \bm f_h(\bm k, t) &= A(\bm k, t) \bm{e_1}(\bm k) + B(\bm k, t) \bm{e_2}(\bm k), \nonumber
\end{align}

where the basis vectors $\bm{e_i}(\bm k) \perp \bm k$ for $i=1, 2$ and $\bm{e_1}(\bm k)\cdot \bm{e_2}(\bm k)^\dagger = 0$,  $A$ and $B$ are complex numbers defined below:

\begin{align}
 A(\bm k, t) = \sqrt{F(k)} \, g_A e^{i \alpha(\bm k)} \nonumber \\
 B(\bm k, t) = \sqrt{F(k)} \, g_B e^{i \alpha(\bm k)} \nonumber
\end{align}

In the above equations, $F(k)$ is a normalisation factor, $g_A^2+g_B^2=1$, and $e^{i\alpha(\bm k)}$ is a uniform random phase. More information about the forcing routines used in eDNS2017 can be found in~\cite{yoffe2013investigation, LinkmannMoritzFrederikLeon2016Spim}.

\section{Paralellisation}

One of the problems with DNS is the memory required to hold the contents of the velocity and magnetic fields in every time step. Since working memory is not unlimited, box sizes larger than $32^3$ cannot be modeled without the use of paralellisation. In computer simulations, this means that the problem is split into different pieces and solved by different processors with (some) communication between them. Maximizing the number of processors while minimising communication between them allows for more time efficient numerical simulations. The paralellisation method used in eDNS is to divide the cubic box of size $N^3 = (2^n)^3$ in $2^m$ (where $m\leq n$) ``slices'' and have the $m$ processors take care of each slice, like shown in fig.~\ref{fig4.1}~\cite{yoffe2013investigation}.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.4\textwidth]{img/slices}
\caption{Paralellisation method through horizontal slices of the cubic box~\cite{yoffe2013investigation}.}
\label{fig4.1}
\end{figure}

The eDNS2017 code uses OpenMPI\footnote{OpenMPI is an open-source Message Passing Interface implementation protocol.} to run on ARCHER, a Cray XC30 supercomputer at the University of Edinburgh’s Advanced Computing Facility. When working on ARCHER, parallel jobs are submitted to a queue and, for a predetermined amount of time, run on a large number of nodes. %Each process has access only to its own data, but they need to know information from the neighboring processes. This means that some communication has to occur between processors.

\section{Running the code}

The code written by Yoffe and Linkmann handles an input file to read the parameters of the simulation, \texttt{input.in}. In said file, the time step \texttt{DT} has to be determined by making use of the CFL condition mentioned in section~\ref{sec4.1}, the simulation time \texttt{TIME} is also stated. The fluid's kinematic viscosity, \texttt{VISC}, and the magnetic Prandtl number \texttt{PRANDTL} can also be modified (and with them the magnetic diffusivity can be changed too). The lattice size, \texttt{LATTICE}, can be changed, but only such that $N=2^n$, where $n \in \mathbb{Z}$. Other parameters like the hypoviscosity, hyporesistivity, cross-helicity, forcing through the magnetic channel, that have not been discussed here can also be changed. For the simulations performed, however, these were all kept unchanged.

A sample input file for eDNS2017 is shown in the Appendix~\ref{apx1}. For said input file, the only things that were changed for the simulations presented in this work were: the

@@

\appendix
\chapter{Sample input file for eDNS2017}
\label{apx1}

Sample input file for a run going to \texttt{TIME~~400} with $\nu=0.9$ and $Pr_M=90$ with helical forcing (\texttt{F\_UHELICITY~~1.0}) in wave number $k_f=2$.

\lstinputlisting[language=bash]{input.in}
